Testing Theories of Expectation Formation using Probabilistic Forecasts

### 2019-06-19 paper/Draft.tex
FIRE benchmmark theory first draft.Remains to proofread.

%opening
\title{Testing Theories of Expectation Formation using Probabilistic Answers}
\author{Tao Wang}

@@

\section{Theories}

\subsection{A Unified Framework}

Agent i's $h$-period ahead density forecast of variable $y_i$ at time $t$ is the conditional density of $y_{i,t+h}$ given the information set $I_{i,t}$ available at time $t$.

$$\widehat f_{i,t}(y^i_{t+h}|I_{i,t})$$

Notice here the $y_{t+h}$ has supscript $i$. This is to remind that in general the variables $y_{t+h}$ can be either aggregate variables or idiosyncratic ones, in the former case, we could drop subscript $i$.

$I_{i,t}$ is the information set available at time t for individual $i$. The information set can be also agent specific, therefore, it has subscript $i$.  The specific content contained in $I_t$ varies from different models. Seemingly a simple and justified assumption is that $x_t \in  I_t$. This, however, may not be ture for all models. For instance, sticky expectation and rational inattention literature all predict that agents are not able to update new information instataneously. Thus the information set may not winclude the recent realization of the variable itself.

Assumption also needs to be made about the underlying model that generates the conditional density of the variables by agent $i$.

Given the forecast density, the expectation of forecast at time $t$ is.

$$E_{i,t}(y_{t+h}|I_{i,t}) =\int \widehat f_{i,t}(y_{t+h}|I_{i,t})y_{t+h} d y_{t+h}\equiv \bar y_{i,t,t+h}$$

The forecasting variance is

$$Var_{i,t}(y_{t+h}|I_{i,t})=\int \widehat f_{i,t}(y_{t+h}|I_{i,t}) (y_{t+h} - \bar y_{i,t,t+h})^2d y_{t+h}$$

The true process of $y_{t}$ depends on the variables. Here, we consider a simple case where true process of $y_t$ is AR(1).

@@

Full information rational expectation implies $y_t \in I_t$

\textbf{Expectation}
$$E_{i,t}(y_{t+h}|y_t) = E_{i,t}(y_{t+h}|y_t) = \rho^h y_t $$

\textbf{Variance} 
$$Var_{i,t}(y_{t+h}|y_t) = \sum^{h}_{s=1}\rho^{2s} \sigma^2_{\omega}$$

Ex post forecast error of individual i $$FE_{i,t} = y_{t+h} - E_{i,t}(y_{t+h}|y_t) = \sum^h_{s=1} \rho^s \omega_{t+h-s}$$ has is mean zero and orthogonal to information at time $t$.

A special case below.

@@

>>>>

%opening
\title{Testing Theories of Expectation Formation using Probabilistic Forecasts}
\author{Tao Wang}

@@

\section{Theories of Expectation Formation}

\subsection{Definition of moments}

An agent $i$ is forming expectations about a stochastic variable $y^i_{t+h}$. The supscript $i$ can be dropped if it is an aggregate (i.e. inflation) variable instead of individual specific  (i.e. household income or house value). This paper focuses on forecasting of aggregate variables, in particular, inflation. So we can simply denote the variable as $y_{t+h}$. To put it another way, all agents are foreasting the same object. \footnote{Only in the context of forecasting aggregate variable, it is meaningful to study the average expectations and disagreements across agents.}

Denote $ f_{i,t+h|t}$ as agent i's $h$-period-ahead density forecast. $ f_{i,t+h|t}$ is the conditional density of $y_{i,t+h}$ given the information set $I_{i,t}$ available at time $t$.

$$f_{i,t+h|t} \equiv f_{i,t}(y_{t+h}|I_{i,t})$$

$I_{i,t}$ is the information set available for individual $i$ at time t. The information set can be  agent specific, thus it has subscript $i$.  The specific content contained in $I_t$ varies from different models of expectation. For instance, sticky expectation and rational inattention literature all assume that agents are not able to update new information instataneously. So the information set may not contain the most recent realization of the variable of forecast $y_t$. \footnote{Given the same information set available to agents, different theories may also differ in the underlying models each agent use to form the conditional density of the variables by agent $i$. Examples of such include multi-prior or model uncertainty. \cite{xx}. }

Accordingly, h-period ahead mean forecast at $t$, denoted as $ y_{i,t+h|t}$, is the conditional expectation of $y_{t+h}$ by the agent $i$.

$$y_{i,t+h|t} \equiv E_{i,t}(y_{t+h}) =\int f_{i, t+h|t} d y_{t+h}$$

Similarly, individual forecasting variance $\sigma_{i,t+h|t}$, hereafter termed as indiviual uncertainty in this paper, is the conditional variance.

$$\sigma^2_{i,t+h|t} \equiv Var_{i,t}( y_{t+h} )$$

Individual forecast error $FE_{i,t+h|t}$ is the difference of ex post realized value of $y_{t+h}$ and individual foreacast at time $t$.

$$FE_{i,t+h|t} = y_{i,t+h|t} - y_{t+h}$$

The population analogues of individual mean forecast, uncertainty and forecast errors are simply the average of the individual moments taken across agents. Denote them as $\bar y_{t+h|t}$, $\bar \sigma^2_{t+h|t}$, and $\overline{FE}_{t+h|t}$, respectively. Hereafter, they are termed as population mean forecast, population uncertainty and population forecast error, respectively. In addition, disagreements is defined as the cross-sectional variance of mean forecasts of individual agents.  Denote it as $\overline{Var}_{t+h|t}(y_{i,t+h|t}) $. To simplify notation, let us directly call it $\overline{Disg}_{t+h|t}$.  \footnote{This is the same termiology used by \cite{xx}.}

In summary, we have 3 individual moments and 4 population moments listed in below.

\begin{table}[]
	\centering
		\caption{Definition and Notation of Moments}
	\begin{tabular}{ll}

		\hline 
		Individual Moments                                  & Population Moments                             \\
		\hline 
	Mean orecast: $y_{i,t+h|t}$                   & Average forecast: $\bar y_{t+h|t}$                   \\
		Forecast error: $FE_{i,t+h|t}$ & Average forecast error: $\overline{FE}_{t+h|t}$ \\
		Uncertainty: $\sigma^2_{i,t+h|t}$         & Average uncertainty:  $\bar \sigma^2_{i,t+h|t}$ \\
		& Disagreements:  $\overline{Disg}_{t+h|t}$       \\
		\hline 
	\end{tabular}
\end{table}

Finally, we assume the underlying true process of $y_{t}$ is $AR(1)$ with persistency parameter $0<\rho <1$ and i.i.d. shock $\omega_t$.

@@

In the full information rational expectation (FIRE) benchmark,  we assume all agents perfectly observe $y_t$ at time $t$ and understand the true process of $y$. Therefore, individual forecast is $\rho^h y_t $. This is also shared by all agents. Therefore, it is also equal to average forecast.

Both individual and population forecast errors are simply the realized shocks between $t+1$ to $t+h$.

$$\overline{FE}_{i,t+h|t} = \sum^{h-1}_{s=1} \rho^s \omega_{t+h-s}$$ It is easy to see that the forecast error is orthogonal to information available till time $t$. This provides the a well known null hypothesis of FIRE. \cite{xx}

The second testable null hypothesis of FIRE here is that forecast errors at two points of time $h$ or more periods apart are not serially correlated. This is not the case within $h$ periods as the realized shocks in overlapping periods enter both forecast errors.

$$Cov(\overline{FE}_{t+h|t}, \overline{FE}_{t+s+h|t+s}) = 0 \quad \forall s \geq h$$

The predictions holds at both individual and population levels.

Another hypothesis test of FIRE relies upon disagreements. \footnote{For instance, \cite{mankiw2003disagreement} documents substantial time-varying disagreements of inflation uses consumers and professional forecasters' survey. } As agents perfect update the same information, there is no disagreements at any point of the time.

$$\overline{Disg}_{t+h|t}=0 \quad \forall t$$

A corollary is that disagreements do not respond to realized shocks any any point of the time.  Empricial tests of this kind include \cite{xx}.\cite{xx} etc.

In addition to forecast error and disagreements, FIRE also predicts that all individual shares the same degree of uncertainty. The same degree of uncertainty simply comes from uncertainty about future unrealized shocks between $t$ and $t+h$. In \cite{}'s termology, not only all agents agree about the mean forecast, but also agree about the uncertainty according to FIRE.

$$\bar \sigma^2_{t+h|t} = \sum^{h}_{s=1}\rho^{2s} \sigma^2_{\omega}$$

The time series pattern of uncertainty according to FIRE is less obvious. Weather it is time variate depends on if the variance of the shocks $\omega_t$ is time varying. Although in our baseline case we make such an assumption, in general it may not be true.

It turns out, however, the serial correlation of uncertainty across different horizons of forecasting does follow testable pattern according to FIRE. In particular, consistent with our current notation, let us denote the forecast of $y_{t+h}$ over horizon of $h-k$ as $y_{t+h|t+k} \quad \forall k =0,1...h$.

$$\bar \sigma^2_{t+h|t+k} = \sum^{h-k}_{s=1}\rho^{2s} \sigma^2_{\omega}$$

As the forecasting horizon decreases relative to a given future point of time, there is a pure efficiency gain or reduction in uncertainty as more and more shocks have realized. So the uncertainty should drop unambigously.

Quanlitatively, $\bar \sigma^2_{t+h|t+k}$ should be smaller than $\bar \sigma^2_{t+h|t}$ $\forall 1<k\leq h$.

Compared to other null hypotheses discussed above, this is a weaker test. It is true that rational forecasting relies upon utilizing all the information available at any point of time in the most efficient way.

@@

*****************************************************
